---
title: "Exploratory - clustering - Stamp 3 - c20k"
subtitle: "Emanuele Pitino"
date: "`r Sys.Date()`"
format: 
  html:
    theme: superhero
    smooth-scroll: true
    code-fold: true
    self-contained: true
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
editor: source
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  fig-width: 16     
  fig-height: 12 
---

### Libraries

```{r}
suppressPackageStartupMessages({
library(SingleCellExperiment)
library(dplyr)
library(here)
library(scater)
library(glue)
library(qs)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(data.table)
library(InSituType)
})
```

### Data loading

```{r}
dir <- glue("{here()}")
source(glue("{dir}/scripts/misc/paths.R"))
source(glue("{dir}/scripts/misc/BIN.R"))
```

```{r}
stamp <- "stamp_3"
sample <- "c20k"
dir <- glue("{here()}")
source(glue("{dir}/scripts/misc/paths.R"))
source(glue("{dir}/scripts/misc/BIN.R"))

res_dir <- glue("{proj_dir}/data/{stamp}/processed/{sample}")
sce <- qread(glue("{res_dir}/qc_{sample}.qs"), nthreads = 8)
#sce <- sce[,sample(colnames(sce),1000)]
sce 
```

```{r}
dir <- glue("{proj_dir}/data/{stamp}/{sample}/Ist")
dir.create(dir, showWarnings = F, recursive = T)
unsup <- qread(glue("{dir}/unsup.qs"))
```


```{r}
sce$label <- unsup$clust[colnames(sce)]
sce$label_prob <- unsup$prob[colnames(sce)]

pal <- Polychrome::createPalette(26,c("#99FFFF", "#FF99FF", "#FFFF99"))
names(pal) <- unique(sce$label)
```

## Remove cells with transfer prob < 0.8
```{r}
sce <- sce[, sce$label_prob > 0.8]
```

## Cell numbers

```{r}
df <- as.data.frame(table(sce$label)) %>%
  mutate(Proportion = round((Freq / sum(Freq)) * 100, 2))

gg_cnumb <- ggplot(df, aes(x = Var1, y = Freq)) + 
  geom_col(aes(fill = Var1)) +
  geom_text(aes(label = glue("{round(Freq/sum(Freq)*100,1)}%")), vjust = -0.5, size = 5) +
  scale_fill_manual(values = pal) + 
  theme_bw() +
  theme(text = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 15, color = "black"),
        panel.grid = element_blank()) +
  labs(fill = "Sample", y = "# Cells") +
  labs(x = "Cluster", fill = "Cluster")

gg_cnumb
```

## Space
```{r}
df <- as.data.frame(colData(sce))
df <- df[sample(rownames(df)),]

gg_space <-  ggplot(df, aes(x = CenterX_global_px, y = CenterY_global_px, color = label)) + 
  scale_color_manual(values = pal) +
  ggrastr::rasterise(geom_point(size = 0.1, shape = 16), dpi = 600) +
  theme_bw() + 
  theme(panel.grid = element_blank(), text = element_text(colour = "black", size = 20), 
        axis.text = element_text(colour = "black", size = 15)) +
  labs(x = "x_px", y = "y_px") +
  guides(color = guide_legend(override.aes = list(size = 4))) +
   coord_equal() +
   labs(color = "Cluster")

gg_space
```


## FlightPath
```{r}
fp_layout <- flightpath_layout(logliks = unsup$logliks, profiles = unsup$profiles)

gg_fp <- ggplot(fp_layout$cellpos, aes(x = x, y = y, color = fp_layout$clust)) +
  ggrastr::rasterise(geom_point(size = 0.1, shape = 16), dpi = 800) +
#  scale_color_manual(values = pal) +
  theme_bw() +
    theme(text = element_text(size = 15, color = "black"),
          axis.text = element_text(size = 10, color = "black"),
          panel.grid = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(color = "Cluster", title = glue("FlightPath - {stamp} - {sample}")) +
  guides(color = guide_legend(override.aes = list(size = 4))) 

gg_fp
```


## InSituType Profiles

Normalize & log-transform the profiles
```{r}
norm <- unsup$profiles /  Matrix::colSums(unsup$profiles) # normalize
norm <- norm + 1e-6
#lognorm <- log1p(norm) + 1e-6  # log transform
```

Calculate logFC

```{r}
result <- norm 
for (i in 1:nrow(norm)) {
  for (j in 1:ncol(norm)) {
    result[i, j] <- log2(norm[i, j] / mean(norm[i, -j]))
  }
}
```

Take first 10 features for each cluster by logFC

```{r}
feat <- unique(as.vector(unlist(apply(result, 2, function(column) {
  names(sort(column, decreasing = TRUE))[1:10]
}))))
```

Remove outliers at 2.5 sd for better visualization

```{r}
x <- result[feat,]

z <- \(x, th=2.5) {
    if (is.null(dim(x))) {
        x[x < 0] <- 0
        sd <- sd(x, na.rm=TRUE)
        x <- x-mean(x, na.rm=TRUE)
        if (sd != 0) x <- x/sd
    } else {
        mus <- colMeans(x)
        sds <- colSds(x)
        x <- sweep(x, 2, mus, `-`)
        x <- sweep(x, 2, sds, `/`)
    }
    x[x > +th] <- +th
    x[x < -th] <- -th
    return(x)
}
mtx_scaled <- z(x)
```

Plot

```{r}
# Perform hierarchical clustering on euclidean distance matrix with "average" method
column_dist_euclidean <- dist(mtx_scaled, method = "euclidean")
column_hclust_complete <- hclust(column_dist_euclidean, method = "average")
#colnames(mtx_scaled) <- c("LnCAP", "SKBR3", "MCF7")

# Create the heatmap
color_function <- circlize::colorRamp2(c(min(mtx_scaled), 0, max(mtx_scaled)), c("navy", "white", "red"))
hm <- ComplexHeatmap::Heatmap(t(mtx_scaled),
              name = "expression",
              cluster_rows = F,
              cluster_columns = as.dendrogram(column_hclust_complete),
              show_row_names = TRUE,
              show_column_names = TRUE,
              show_row_dend = FALSE,
              show_column_dend = FALSE,
              col = color_function,
              row_names_side = "left",
              height = unit(4,"cm"),
              column_names_gp = grid::gpar(fontsize = 8))

hm
```

```{r}
sce$label[sce$label == "a"] <- "SKBR3"
sce$label[sce$label == "b"] <- "MCF7"
sce$label[sce$label == "c"] <- "LnCAP"
```

```{r}
pdf(glue("{plt_dir}/{stamp}/{sample}/a.pdf"))
gg_cnumb
dev.off()

pdf(glue("{plt_dir}/{stamp}/{sample}/b.pdf"))
gg_fp
dev.off()

pdf(glue("{plt_dir}/{stamp}/{sample}/c.pdf"))
gg_space
dev.off()

pdf(glue("{plt_dir}/{stamp}/{sample}/e.pdf"))
hm
dev.off()
```

```{r}
dir <- glue("{proj_dir}/data/{stamp}/processed/combined")
dir.create(dir, showWarnings = F, recursive = T)
cd <- as.data.frame(colData(sce))
qsave(cd,glue("{dir}/cd_{sample}.qs"))
```
