---
title: "First annotation round - Stamp3 PBMCs"
subtitle: "Emanuele Pitino"
date: "`r Sys.Date()`"
format: 
  html:
    theme: superhero
    smooth-scroll: true
    code-fold: true
    self-contained: true
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
editor: source
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  fig-width: 16     
  fig-height: 12 
---
In this script we will remove any low-quality or artifact clusters in this stamp, which will be re-processed and re-clustered.

### Libraries

```{r}
suppressPackageStartupMessages({
library(SingleCellExperiment)
library(dplyr)
library(here)
library(scater)
library(glue)
library(qs)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(data.table)
library(spatstat)
library(InSituType)
})
```

### Data loading

```{r}
dir <- glue("{here()}")
source(glue("{dir}/scripts/misc/paths.R"))
source(glue("{dir}/scripts/misc/BIN.R"))
```

```{r}
stamp <- "stamp_3"
sample <- "PBMCs"
res_dir <- glue("{proj_dir}/data/{stamp}/{sample}/Ist")
sce <- qread(glue("{res_dir}/anno_sce.qs"), nthreads = 8)
```

## Palette
```{r}
pal <- Polychrome::createPalette(26,c("#99FFFF", "#FF99FF", "#FFFF99"))
names(pal) <- unique(sce$label)
```

## Viz
We see cluster e to have a precise spatial localization on the top-right corner of the slide. This might be a technical artifact during wet-lab processing. Also, from the Heatmap in Exploratory_Ist.html, we can see that its profile does not mach any known PBMCs subtype.
```{r}
df <- as.data.frame(colData(sce))
df <- df[sample(rownames(df)),]

gg_space <-  ggplot(df, aes(x = CenterX_global_px, y = CenterY_global_px, color = label)) + 
  scale_color_manual(values = pal) +
  ggrastr::rasterise(geom_point(size = 0.1, shape = 16), dpi = 600) +
  theme_bw() + 
  theme(panel.grid = element_blank(), text = element_text(colour = "black", size = 20), 
        axis.text = element_text(colour = "black", size = 15)) +
  labs(x = "x_px", y = "y_px") +
  guides(color = guide_legend(override.aes = list(size = 4))) +
   coord_equal() +
   labs(color = "Cluster")

gg_space
```

### Filter
```{r}
sce <- sce[,sce$label != "e"]
```

## Save
```{r}
dir <- glue("{proj_dir}/data/{stamp}/{sample}/Ist")
qsave(sce, file = glue("{dir}/clean_sce.qs"), nthreads = 8)
```
