---
title: "PreProcessing - Flex iPSC - Stamp 7b - Combined"
subtitle: "Emanuele Pitino"
date: "`r Sys.Date()`"
format: 
  html:
    theme: superhero
    smooth-scroll: true
    code-fold: true
    self-contained: true
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
editor: source
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  fig-width: 9     
  fig-height: 6 
---
This document provides a comprehensive overview of the preprocessing steps performed on the stamp_7b iPSCs Flex dataset.

Inputs are Sce objects that were previously quality-controlled separately (outliers on quality metrics + doublets flag). 
The analysis covers the following:
1. Normalization; 2. Feature selection; 3. Dimensionality reduction, PCA followed by UMAP.

## Setup
### Libraries
```{r}
suppressPackageStartupMessages({
library(SingleCellExperiment)
library(dplyr)
library(here)
library(scater)
library(scuttle)
library(scran)
library(data.table)
library(glue)
library(qs)
library(scales)
})
```

### Paths
```{r}
dir <- glue("{here()}")
source(glue("{dir}/scripts/misc/paths.R"))
source(glue("{dir}/scripts/misc/BIN.R"))

sample <- "combined"
stamp <- "Stamp_7b"
#npcs <- params$npcs
```

### Load data
```{r}
dirs <- list.dirs(glue("{proj_dir}/data/{stamp}/processed/flex")) # get dirs
dirs <- dirs[grepl("^iPSC_|(mesoderm|ectoderm|endoderm)$", basename(dirs))]
sce <- SingleCellExperiment() # init sce obj

sce <- list()

sce_list <- lapply(dirs, function(dir) {
  qread(glue("{dir}/clust_sce.qs"), nthreads = 8)
})

# Apply modifications to each element in sce_list
sce_list <- lapply(sce_list, function(tmp) {
  reducedDims(tmp)$PCA <- NULL                # Remove PCA
  reducedDims(tmp)$UMAP <- NULL               # Remove UMAP
  assays(tmp)$logcounts <- NULL               # Remove logcounts
  colnames(tmp) <- paste0(tmp$sample, "_", colnames(tmp))  # Rename barcodes
  rowData(tmp)$scDblFinder.selected <- NULL   # Remove scDblFinder stats from rowData
  metadata(tmp)$scDblFinder.stats <- NULL     # Remove scDblFinder stats from metadata
  metadata(tmp)$scDblFinder.threshold <- NULL # Remove scDblFinder threshold from metadata
  tmp  # Return the modified object
})

sce <- do.call(cbind,sce_list)

rm(sce_list)
gc()
```

# Processing

## LogNorm
```{r}
#sce <- sce[,sample(colnames(sce),1000)]
sce <- sce[,sce$sample != "endoderm"]
sce <- logNormCounts(sce)
```

## Feature Selection

Features were selected independently for each library. In this step, we will compute the intersection of these features across all libraries and use them in downstream analyses.
```{r}
#hvg_list <- metadata(sce)[grep("hvg_", names(metadata(sce)), value = TRUE)]
#hvg_list <- lapply(hvg_list, function(x) x[1:2000])
#hvg <- unique(unlist(hvg_list))
#metadata(sce)$hvg_combined <- hvg

#hvg_list <- metadata(sce)[grep("hvg_", names(metadata(sce)), value = TRUE)]
#hvg <- sapply(hvg_list, \(.) rownames(sce) %in% .)
#sum(rowSums(hvg) > 2)

set.seed(0010101)
dec.var <- modelGeneVar(sce, BPPARAM = bp) 
hvg <- getTopHVGs(dec.var, fdr.threshold = 1)
length(hvg)
dec.var$hvg <- "no" # Assign to dec.var column for plot
dec.var$hvg[rownames(dec.var) %in% hvg] <- "yes"
gg_hvg <- plot_hvg(dec.var = dec.var, sub = sample) # plot
gg_hvg
```

## PCA on the whole panel

```{r}
set.seed(101001)
sce <- fixedPCA(sce, subset.row = NULL)

num_pcs_to_retain <- 10
percent.var <- attr(reducedDim(sce), "percentVar")

# Create a data frame for ggplot
data <- data.frame(PC = 1:length(percent.var), Variance = percent.var)
# Plot
gg_var <- ggplot(data, aes(x = PC, y = Variance)) +
  geom_point() +
  xlab("PC") +
  ylab("Variance explained (%)") +
  geom_vline(xintercept = num_pcs_to_retain, color = "red") +
  theme_bw()
gg_var
```

```{r}
reducedDim(sce, "PCA") <-  reducedDim(sce, "PCA")[,1:num_pcs_to_retain]
wh(6,5)
gg_pca <- plotPCA(sce, scattermore = TRUE, point_size = 2) + ggtitle("PCA")
gg_pca
```

## Run UMAP
```{r}
set.seed(123)
sce <- runUMAP(sce, dimred="PCA", BPPARAM = bp)
gg_um <- plotReducedDim(sce, "UMAP", scattermore = TRUE, point_size = 2) 
gg_um
```


## Add Library information
```{r}
sce$library <- "B3"
```

## PCA - time points

```{r}
df <- as.data.frame(colData(sce)) # cd
pca <- reducedDim(sce,"PCA") # pc
df <- cbind(df,pca) # merge
df <- df[sample(nrow(df)), ] # shuffle

pal <- Polychrome::createPalette(n_distinct(sce$sample), c("#FBB4AE", "#B3CDE3", "#CCEBC5"))
names(pal) <- unique(sce$sample)

df$sample <- factor(df$sample, levels = c("iPSC_parental","mesoderm","ectoderm"))

ggplot(df, aes(x = PC1, y = PC2, color = sample)) +
  ggrastr::rasterise(geom_point(shape = 16, size = 0.1), dpi = 1000) +
  theme_bw() +
  scale_color_manual(values = pal) +
  theme(
    text = element_text(size = 25, color = "black"),
    axis.text = element_text(size = 20, color = "black"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 18),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(x = "PC1", y = "PC2", color = "Sample") +
  coord_equal()
```

# Clustering on un-integrated PCA

## Build snn graph
```{r}
snn <- buildSNNGraph(sce, type = "jaccard", use.dimred="PCA", BPPARAM = bp)
```

## Run louvain 
```{r}
clusters <- igraph::cluster_louvain(snn, resolution = 0.5)
```

```{r}
sce$label <- as.character(clusters$membership) # assign labels
pal <- Polychrome::createPalette(n_distinct(sce$label), c("#FBB4AE", "#B3CDE3", "#CCEBC5"))
names(pal) <- unique(sce$label)
```

## Clusters numbers

```{r, fig.width= 12, fig.height= 6}
df <- as.data.frame(colData(sce))
df <- as.data.frame(table(df$label))  %>%
        mutate(pct = round(Freq/sum(Freq),2))
df$Var1 <- factor(df$Var1, levels = df$Var1[order(-df$Freq)])

ggplot(df, aes(x = Var1, y = Freq, fill = Var1)) + 
  geom_col() + 
  geom_text(aes(label = Freq), vjust = -0.5, size = 3) +
  scale_fill_manual(values = pal) + 
  theme_bw() +
  theme(text = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 10, color = "black")) +
  labs(x = "Cluster", y = "nCells", fill = "Cluster")
```

## Clusters proportions per sample
```{r, fig.width= 12, fig.height= 6}
df <- as.data.frame(table(sce$sample,sce$label)) %>%
      group_by(Var1) %>%
      mutate(Proportion = round((Freq / sum(Freq)) * 100, 2))

df$Var1 <- factor(df$Var1, levels = c("iPSC_parental","mesoderm","ectoderm"))
ggplot(df, aes( x = Var1 , y = Proportion, fill = Var2)) + 
  geom_col() +
  scale_fill_manual(values = pal) +
  theme_bw() +
  theme(text = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 15, color = "black"),
        panel.grid = element_blank()) +
  labs(x = "Sample", fill = "Cluster")
```

## QC metrics per cluster

```{r, fig.width= 12, fig.height= 6}
df <- as.data.frame(colData(sce))

# reorder by mean ncount
df$label <- factor(df$label, levels = df %>% 
  group_by(label) %>% 
  summarise(mean_sum = mean(sum)) %>% 
  arrange(-mean_sum) %>% 
  pull(label))

# Create the plot
ggplot(df, aes(x = label, y = sum)) + 
  geom_jitter(shape = 16, alpha = 0.8, size = 0.8, aes(color = log(detected))) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA, fill = NA) +
  scale_color_viridis_c() +
  scale_y_log10() +
  labs(x = "Cluster", y = "nCount", color = "Log(nFeature)") + 
  theme_bw() +
  theme(text = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 10, color = "black"),
        panel.grid = element_blank())
```


## nFeature/nCount
```{r}
df <- as.data.frame(colData(sce))

ndbl <- sum(df$scDblFinder.class == "doublet")
nsgl <- sum(df$scDblFinder.class == "singlet")

pctdbl <- round(ndbl/(ndbl+nsgl)*100,2)

df <- df[sample(rownames(df)),]
ggplot(df, aes(x = detected, y = sum, color = scDblFinder.class)) +
  geom_point(shape = 16, size = 0.4) +
  scale_color_manual(values = c("singlet" = "green4", "doublet" = "red4")) +
            labs(subtitle = glue("Doublets: {pctdbl}%") ,x = "nFeature", y = "nCount", color = "scDblFinder", title = "nFeat vs nCount") + 
            theme_bw() +
            theme(text = element_text(size = 15, color = "black"),
                  axis.text = element_text(size = 10, color = "black")) +
            guides(color = guide_legend(override.aes = list(size = 3)))
```

## Doublets per cluster
```{r}
df <- as.data.frame(colData(sce))
df <- as.data.frame(table(df$label, df$scDblFinder.class))

# Calculate the percentage and reorder by 'singlet' percentage
df <- df %>%
  group_by(Var1) %>%
  mutate(pct = round(Freq / sum(Freq), 2)) %>%
  ungroup()

# Calculate the percentage of 'singlets' for reordering
singlet_pct <- df %>% 
  filter(Var2 == "singlet") %>% 
  select(Var1, pct)

# Reorder Var1 by the percentage of 'singlets'
df$Var1 <- factor(df$Var1, levels = singlet_pct$Var1[order(singlet_pct$pct, decreasing = TRUE)])

# Make Var2 a factor to control the order in the legend and plot
df$Var2 <- factor(df$Var2, levels = c("doublet", "singlet"))

# Create the plot
ggplot(df, aes(x = Var1, y = pct, fill = Var2)) +
  scale_fill_manual(values = c("singlet" = "green4", "doublet" = "red4")) +
  geom_col(position = "stack") + 
  labs(x = "Cluster", y = "Proportion", fill = "scDblFinder", title = "Doublets % per Cluster") + 
  theme_bw() +
  theme(text = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 10, color = "black")) + 
  guides(color = guide_legend(override.aes = list(size = 3)))
```

# Save plots
```{r}
combined <- wrap_plots(gg_var, gg_hvg, gg_pca, gg_um, ncol = 2, nrow = 2) +
  plot_annotation(tag_levels = "A") + 
  plot_annotation(title = glue("Flex - Stamp 7b - iPSCs - {sample}"), subtitle = glue("N = {comma(ncol(sce))} cells"))

plotdir <- glue("{plt_dir}/{stamp}/flex/iPSC/{sample}")
dir.create(plotdir, showWarnings = F, recursive = T)

pdf(glue("{plotdir}/PreProc_{stamp}_{sample}.pdf"), width = 20, height = 12)
combined
dev.off()
```

# Save data
```{r}
outdir <- glue("{proj_dir}/data/{stamp}/processed/flex/iPSC/{sample}")
dir.create(outdir, showWarnings = F, recursive = T)
qsave(sce, glue("{outdir}/proc_sce.qs"), nthreads = 8)
```
